Sample Clustering

### 2017-02-24 Clustering.Rmd
Updating reports to use with new data

---
title: "Sample Clustering"
author: "Sarah Killcoyne"
date: "12/6/2016"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(GGally)
library(plyr)
library(xlsx)
library(dplyr)
library(pander)
library(Hmisc)
library(ggdendro)
library(reshape2)
library(GenomicRanges)
library(gridExtra)
library(biomaRt)

library(tibble)
library(mclust)

@@

sum.patient.data = as.data.frame(patient.info %>%
  group_by(Patient, Status) %>%
  summarise(
    years=paste(range(Endoscopy.Year), collapse='-'),
    n.samples=length(Samplename), med.cellularity=median(Barretts.Cellularity,na.rm=T),
    mean.reads=mean(Total.Reads),
    sd.reads=sd(Total.Reads)
  ))

pander(sum.patient.data)
```

@@

ttr = t.test(subset(patient.info, Status == 'NP')$Total.Reads, subset(patient.info, Status == 'P')$Total.Reads)
ttrbe = t.test(subset(be, Status == 'NP')$Total.Reads, subset(be, Status == 'P')$Total.Reads)
ttrlgd = t.test(subset(lgd, Status == 'NP')$Total.Reads, subset(lgd, Status == 'P')$Total.Reads)

tbclgd = t.test(subset(lgd, Status == 'NP')$Barretts.Cellularity, subset(lgd, Status == 'P')$Barretts.Cellularity)
tbcbe = t.test(subset(be, Status == 'NP')$Barretts.Cellularity, subset(be, Status == 'P')$Barretts.Cellularity)

t.test(subset(patient.info, Pathology == 'BE')$Barretts.Cellularity, 
               subset(patient.info, Pathology == 'LGD')$Barretts.Cellularity)

t.test(subset(patient.info, Pathology == 'LGD')$Barretts.Cellularity, 
               subset(patient.info, Pathology == 'HGD')$Barretts.Cellularity)

## No differences between the number of reads by pathology
t.test(subset(patient.info, Pathology == 'BE')$Total.Reads, subset(patient.info, Pathology == 'LGD')$Total.Reads)
t.test(subset(patient.info, Pathology == 'BE')$Total.Reads, subset(patient.info, Pathology == 'HGD')$Total.Reads)
t.test(subset(patient.info, Pathology == 'LGD')$Total.Reads, subset(patient.info, Pathology == 'HGD')$Total.Reads)

t.test(subset(patient.info, Pathology == 'BE')$Barretts.Cellularity, subset(patient.info, Pathology == 'LGD')$Barretts.Cellularity)
t.test(subset(patient.info, Pathology == 'BE')$Barretts.Cellularity, subset(patient.info, Pathology == 'HGD')$Barretts.Cellularity)
t.test(subset(patient.info, Pathology == 'LGD')$Barretts.Cellularity, subset(patient.info, Pathology == 'HGD')$Barretts.Cellularity)

There are a few things we need to be careful of and a couple of questions to check here:

1. The total number of reads per sample between NP and P might show a difference, but currently with multi-test correction it's not significant.
`r  signif(ttr$p.value,2)*11`
So are those samples with lower numbers of reads showing less complexity overall?  When sequencing NP are we likely to get fewer reads than with P? If we look only at the samples with BE (no LGD or HGD samples) the difference disappears `r signif(ttrbe$p.value, 2)*11`. If we include LGD with the BE samples we again see the difference.  `r signif(ttrlgd$p.value, 2)*11`.  This is suggestive that changes in copy number result in more reads being output.

2. There is a significant difference in the cellularity of the Barretts sample, with progressors having a higher mean cellularity, even when only comparing samples that were either LGD or BE.  `r signif(tbclgd$p.value, 2)` However, even the BE only samples show a higher cellularity overall `r signif(tbcbe$p.value, 2)*11`  That said, currently we have a high rate of missing information, with about `r length(which(is.na(subset(patient.info, Status == 'P')$Barretts.Cellularity)))/length(which(patient.info$Status == 'P'))*100`% of the cellualarities missing for progressors and `r length(which(is.na(subset(patient.info, Status == 'NP')$Barretts.Cellularity)))/length(which(patient.info$Status == 'NP'))*100`% of the NP cellularities missing.

3. With regards to p53 staining, non-progressors have no p53 positive tests, while progressors had `r p53Status[2,2]` overall.
`r pander(p53Status, justify='left', caption="P53 staining status for progressors vs non")`
Of these `r table(be$p53.Status)[['1']]` were positive in a BE sample, and `r table(lgd$p53.Status)[['1']]` were positive when LGD samples were included.

@@

```{r pca, echo=F, warning=F, message=F, fig.width=20, fig.height=20}
sum.patient.data = arrange(sum.patient.data, Status, n.samples)

@@

    eigenvalues = pca$sdev^2
    screeplot(pca, main=paste(pt, df, 'scree'))

@@

    gg = ggplot(df, aes(PC1, PC2)) + geom_line() +
        geom_label(aes(label=year),nudge_y=0.04) + 
        geom_point( aes(color=Pathology, size=Pathology)) +
        labs(title=paste(pt, sv))
  plots[[sv]] = gg
  }  
  do.call(grid.arrange, c(plots, list(ncol=2))
  return(gg)
}

plot.pca.norm<-function(pt) {

  x1 = patient.data[[pt]]$norm.seg.vals[, patient.data[[pt]]$info$Samplename]
  x1 = x1[,patient.data[[pt]]$info$Samplename]
  colnames(x1) = paste(patient.data[[pt]]$info$Endoscopy.Year, '(', patient.data[[pt]]$info$Pathology, ')', sep='' )

  eigenvalues = pca$sdev^2
  #screeplot(pca, main='scree')

  gg = ggplot(df, aes(PC1, PC2)) + geom_line() +
    geom_label(aes(label=year),nudge_y=0.04) + 
    geom_point( aes(color=Pathology, size=Pathology)) +
    labs(title=pt)

  return(gg)
}

pplots = lapply(subset(sum.patient.data, Status == 'P')$Patient, plot.pca)
grid.arrange(grobs=pplots, ncol=2, top="Progressors")

npplots = lapply(subset(sum.patient.data, Status == 'NP')$Patient, plot.pca)
grid.arrange(grobs=npplots, ncol=2, top="Non-Progressors")

@@

I tried were several mixture models (e.g. EM) but wasn't able to find a pattern that could reliably separate the few mixtures I did see

```{r clust, echo=T, message=T, warning=F, eval=T, fig.height=6, fig.width=6}

llk = vector(mode='numeric', length=nrow(sum.patient.data))
names(llk) = sum.patient.data$Patient
for (pt in sum.patient.data$Patient) {
  print(pt)
  print(ncol(patient.data[[pt]]$seg.vals[,-(1:5)]))

  x1 = as.matrix(patient.data[[pt]]$seg.vals[,-(1:5)])
  x1 = x1[,patient.data[[pt]]$info$Samplename]

  fit <- Mclust(x1, G=3)
  plot(fit, what='classification')
  title(main=paste(pt, ' (',patient.data[[pt]]$info$Status[1], ')', sep=''))

  llk[pt] = fit$loglik

#   clusterPlots <- data.frame(x1 = clusterData$x1, x2 = clusterData$x2, cluster = factor(model$classification, 
#     levels = c(1, 2), labels = c("z = 0", "z = 1")), uncertainity = model$uncertainty, 
#     logUncertainity = log(model$uncertainty))
# clusterPlots.gg <- ggplot(clusterPlots)
# clusterPlots.gg + geom_point(aes(x = x1, y = x2, color = cluster))

  #tryCatch({
  # model = densityMclust(x1, G=3)
  # plotDensityMclustd(model, data=x1,  col='blue', lwd=2, hist.col = "lightblue",  breaks=100)
  #   mod1dr = MclustDR(model, lambda = 0.5)
  #   #summary(mod1dr)
  #   table(mod1dr$mixcomp)
  # 
  #   #plot(mod1dr, what = "evalues")
  #   #title(main=pt)
  #   plot(mod1dr, what = "density")
  #   title(main=paste(pt, ' (',patient.data[[pt]]$info$Status[1], ')', sep=''))
  # })
}

# Log liklihood doesn't separate them
t.test(llk[subset(sum.patient.data, Status == 'NP')$Patient], llk[subset(sum.patient.data, Status == 'P')$Patient])

```{r means, echo=F, message=F, warning=F, eval=F} 
## Kmeans
#   x1 = as.matrix(patient.data[[pt]]$seg.vals[,-(1:5)])
# 
   fit = kmeans(x1, 3)
#   plot(x1[,c(1,10)], col=km$cluster)
# title(main=paste(pt, ' (',patient.data[[pt]]$info$Status[1], ')', sep=''))

library(cluster) 
clusplot(x1, fit$cluster, color=TRUE, shade=TRUE, labels=2, lines=0, main=pt)

@@

## Using the segmentation values
sim = sum.patient.data[,c('Status','years')]
rownames(sim) = sum.patient.data[,'Patient']
for (pt in sum.patient.data$Patient) {
  patient.data[[pt]]$info = arrange(patient.data[[pt]]$info, Endoscopy.Year, desc(Pathology))

@@

  sim[pt, 'ns'] = net.similarity(ac)
  sim[pt, 'ss'] = sum.similarity(ac)
  sim[pt, 'clusters'] = length(clusters(ac))

@@

sm = as.data.frame(summarise(group_by(sim, Status), mean.ns=mean(ns), sd.ns=sd(ns), mean.ss=mean(ss), sd.ss=sd(ss), mean.c=mean(clusters), sd.c=sd(clusters) ))

@@

fit = glm(Status~clusters, sim, family=binomial(link='logit'))
pander(fit, justify='left')

sim = sim[, c('Status', 'clusters', 'ns','ss')]
actualMatrix = matrix(0,1000,2,dimnames=list(c(), c('prec','recall')))
for (i in 1:nrow(actualMatrix)) {
  trainingRows = sample(1:nrow(sim), nrow(sim)-5)
  train = sim[trainingRows,]
  test = sim[-trainingRows,]

  fit = glm( (as.integer(Status)-1)~clusters, train, family=binomial(link='logit'))
  pred = predict(fit, test, se.fit=F) # probabilities -- of which outcome though?

  actualMatrix[i,'prec'] = getPrecision( actual, predicted )
  actualMatrix[i, 'recall'] = getRecall( actual, predicted )
}

ggplot(as.data.frame(actualMatrix), aes(x=recall, y=prec)) + geom_point() + 
  geom_hline(yintercept = 0.5, col='red', alpha=0.5) +
  labs(title='Precision/Recall for Progressor/NP prediction')

@@

simX = simX[, c('Status', 'clusters', 'ns','ss','gain','loss')]
noHGDPR = as.data.frame(matrix(0,1000,2,dimnames=list(c(), c('prec','recall'))))
for (i in 1:nrow(actualMatrix)) {
  trainingRows = sample(1:nrow(simX), nrow(simX)-5)
  train = simX[trainingRows,]
  test = simX[-trainingRows,]

@@

ggplot(as.data.frame(noHGDPR), aes(x=recall, y=prec)) + geom_point() + 
  geom_hline(yintercept = 0.5, col='red', alpha=0.5) +
  geom_vline(xintercept = 0.5, col='red', alpha=0.5) +
  labs(title='Precision/Recall for Progressor/NP prediction NO HGD')

@@

ggplot(m, aes(x=qv, y=value, col=variable)) + geom_point() +  
  geom_errorbar(aes(ymin=value-sd, ymax=value+sd), width=.01, alpha=0.3) +
  geom_hline(yintercept=5, alpha=0.3) + geom_hline(yintercept=15, alpha=0.3, col='red') + 
  labs(title='Clusters vs q parameters', x='q parameter', y='mean(clusters)')

ggplot(melt(pvals, measure.vars=c('p.clusters'), id.vars='qv'),aes(x=qv,y=value, color=variable)) + geom_point() +
  labs(title='P vs NP p-values per q parameters', y='pvalue NP vs P', x='q param')

>>>>

---
title: "Sample Clustering"
author: "Sarah Killcoyne"
date: "23/02/2016"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(GGally)
library(plyr)
library(xlsx)
library(dplyr)
library(pander)
library(Hmisc)
library(ggdendro)
library(reshape2)
library(GenomicRanges)
library(gridExtra)
library(biomaRt)
library(multtest)
library(tibble)
library(mclust)

@@

sum.patient.data = summarise.patient.info(patient.info)
sum.patient.data = as.data.frame(subset(sum.patient.data, Initial.Analysis)) ## For now
pander(sum.patient.data)
```

@@

set.seed(99)
mtp1 = MTP(X=t(patient.info[,c('Barretts.Cellularity','Total.Reads')]), Y=as.integer((as.integer(patient.info$Pathology) > 3)), robust=T, get.adjp=T, na.rm=T)
mtp2 = MTP(X=t(patient.info[,c('Barretts.Cellularity','Total.Reads')]), Y=patient.info$Status, robust=T, get.adjp=T, na.rm=T)
```

Currently with multi test correction the samples for which cellularity and #of reads are available show no difference for either status (P vs NP)
`r pander(summary(mtp1), justify='left')`

or for samples that are LGD/IMC/HGD vs BE
`r pander(summary(mtp1), justify='left')`

With regards to p53 staining, non-progressors have no p53 positive tests, while progressors had `r p53Status[2,2]` overall.
`r pander(p53Status, justify='left', caption="P53 staining status for progressors vs non")`
Of these `r table(be$p53.Status)[['1']]` were positive in a BE sample, and `r table(lgd$p53.Status)[['1']]` were positive when LGD samples were included.

@@

```{r pca, echo=F, warning=F, message=F, fig.width=20, fig.height=40}
sum.patient.data = arrange(sum.patient.data, Status, total.samples)

@@

    eigenvalues = pca$sdev^2
    screeplot(pca, main=paste(pt, sv, 'scree'))

@@

    gg = ggplot(df, aes(PC1, PC2)) + geom_line() +
        geom_label(aes(label=year),nudge_y=0.04) + 
        geom_point( aes(color=Pathology, size=Pathology)) +
        labs(title=paste(pt, sv))
  plots[[sv]] = gg
  }  
  ga = do.call(grid.arrange, c(plots, list(ncol=2)))
  return(ga)
}

pplots = lapply(subset(sum.patient.data, Status == 'P')$Patient, plot.pca)
grid.arrange(grobs=pplots, nrow=length(subset(sum.patient.data, Status == 'P')$Patient), top="Progressors")

npplots = lapply(subset(sum.patient.data, Status == 'NP')$Patient, plot.pca)
grid.arrange(grobs=npplots, nrow=length(subset(sum.patient.data, Status == 'NP')$Patient), top="Non-Progressors")

@@

I tried were several mixture models (e.g. EM, kmeans) but wasn't able to find a pattern that could reliably separate the few mixtures I did see

### Raw segmented values

#### EM
```{r clust, echo=T, message=T, warning=F, eval=T, fig.height=4, fig.width=4, eval=T}

em.dr<-function(patient, status, x1, var.regions) {
  clusters=c(2:4)
  #llk = matrix(ncol=length(clusters),nrow=nrow(sum.patient.data), dimnames=list(sum.patient.data$Patient, clusters))

  llk = vector(length=length(clusters), mode='numeric')
  names(llk) = as.character(clusters)

## Ok, this doesn't separate regions well either normalized or raw segment values.  Typically the clusters are highly overlapped and not very descriptive
#for (pt in sum.patient.data$Patient) {
    print(pt)
#    status = subset(sum.patient.data, Patient == pt)$Status
#    print(ncol(patient.data[[pt]]$seg.vals[,-(1:5)]))
#    x1 = as.matrix(patient.data[[pt]]$seg.vals[,-(1:5)])
#    x1 = x1[,patient.data[[pt]]$info$Samplename]
    #x1 = as.matrix(patient.data[[pt]]$norm.seg.vals)

    for (g in clusters) {
      print(g)
      model = densityMclust(x1, G=g)
      highCluster = which.max(colSums(model$parameters$mean))

      tryCatch({
        mod1dr = MclustDR(model, lambda = 0.5)

        #     par(mfrow=c(2,1)) 
        plot(mod1dr, what="evalues")
        title(paste(pt, '(',status,') G=', g, sep=''))
        plot(mod1dr, what='classification') # while this looks like PCA, it's not quite
        title(paste(pt, '(',status,') G=', g, sep=''))
        #     dev.off()
        plot(mod1dr, what='density', dimens=1)
        title(paste(pt, '  G=', g, sep=''))
      }, error = function(e) {
        cat(paste(pt, g, "failed DR"))
      })

      a1 = rownames(x1[which(model$classification == highCluster),] )
      if (nrow(var.regions) > 0) {
        am = merge(a1, var.regions, all=T)
        pander( as.data.frame(rownames(x1[which(model$classification == highCluster),])),
             caption=paste("From the cluster with the highest means the regions above are selected, of these ",length(setdiff(rownames(var.regions), a1)),
                           " regions are not found in the variable regions selected using cutoffs.", length(intersect(rownames(var.regions), a1)), " are in both."))
      } else {
        pander(rownames(x1[which(model$classification == highCluster),]),
             caption=paste("From the cluster with the highest means the regions above are selected. This patient had no variable regions selected based on cutoffs"))

      }

      llk[as.character(g)] = model$loglik
    }
    return(llk)
}

llk = as.data.frame(matrix(nrow=nrow(sum.patient.data),ncol=3,dimnames=list(c(sum.patient.data$Patient), c(2:4))))
for (pt in sum.patient.data$Patient) {
    print(pt)
    status = subset(sum.patient.data, Patient == pt)$Status

    x1 = patient.data[[pt]]$seg.vals
    rownames(x1) = apply(x1, 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

    var = patient.data[[pt]]$variable.regions
    if (nrow(var) > 0)
      rownames(var) = apply(var, 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep='')))

    llk[pt,] =  em.dr(pt,status,as.matrix(x1), var[,-(1:5)])
  }

ggplot(melt(as.matrix(llk)), aes(x=Var1, y=value, color=Var1)) + geom_point() + 
  theme(axis.text.x=element_text(angle = 45), legend.position='none')
  labs(title="Raw segment value log liklihoods", x='', y='loglik')

# Log liklihood doesn't separate them individually or the range
llk.diff = apply(llk, 1, function(x) diff(range(x)) )
wilcox.test(llk.diff[subset(sum.patient.data, Status == 'NP')$Patient], llk.diff[subset(sum.patient.data, Status == 'P')$Patient])

#t.test(llk[subset(sum.patient.data, Status == 'NP')$Patient,'2'], llk[subset(sum.patient.data, Status == 'P')$Patient,'2'])
#t.test(llk[subset(sum.patient.data, Status == 'NP')$Patient,'3'], llk[subset(sum.patient.data, Status == 'P')$Patient,'3'])
#t.test(llk[subset(sum.patient.data, Status == 'NP')$Patient,'4'], llk[subset(sum.patient.data, Status == 'P')$Patient,'4'])
```

#### Kmeans
```{r kmeansraw, echo=F, message=F, warning=F, eval=F, fig.height=6, fig.width=6} 
## Kmeans
#   x1 = as.matrix(patient.data[[pt]]$seg.vals[,-(1:5)])
# 
for (pt in sum.patient.data$Patient) {
    print(pt)
    status = subset(sum.patient.data, Patient == pt)$Status
    x1 = patient.data[[pt]]$seg.vals
    rownames(x1) = apply(patient.data[[pt]]$seg.vals[,c(1:4)], 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

    fit = kmeans(x1, 3)
    autoplot(fit, data=x1, label=T) + labs(title=paste(pt, '(', status, ') raw seg', sep=''))
    pander(patient.data[[pt]]$variable.regions[,(1:4)], caption='Variable regions selected by cutoffs')
}
```

### Normalized segmented values
```{r clust2, echo=T, message=T, warning=F, eval=T, fig.height=4, fig.width=4, eval=T}
llk = as.data.frame(matrix(nrow=nrow(sum.patient.data),ncol=3,dimnames=list(c(sum.patient.data$Patient), c(2:4))))
for (pt in sum.patient.data$Patient) {
    print(pt)
    status = subset(sum.patient.data, Patient == pt)$Status
    x1 = patient.data[[pt]]$norm.seg.vals
    rownames(x1) = apply(patient.data[[pt]]$seg.vals[,c(1:4)], 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

    var = patient.data[[pt]]$variable.regions
    if (nrow(var) > 0)
      rownames(var) = apply(var, 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep='')))

    llk[pt,] = em.dr(pt,status,x1, var)
  }

ggplot(melt(as.matrix(llk)), aes(x=Var1, y=value, color=Var1)) + geom_point() + 
  theme(axis.text.x=element_text(angle = 45), legend.position='none') + 
  labs(title="Normalized segment value log liklihoods", x='', y='loglik')

# Log liklihood doesn't separate them individually or the range
llk.diff = apply(llk, 1, function(x) diff(range(x)) )
wilcox.test(llk.diff[subset(sum.patient.data, Status == 'NP')$Patient], llk.diff[subset(sum.patient.data, Status == 'P')$Patient])

#t.test(llk[subset(sum.patient.data, Status == 'NP')$Patient,'2'], llk[subset(sum.patient.data, Status == 'P')$Patient,'2'])
#t.test(llk[subset(sum.patient.data, Status == 'NP')$Patient,'3'], llk[subset(sum.patient.data, Status == 'P')$Patient,'3'])
#t.test(llk[subset(sum.patient.data, Status == 'NP')$Patient,'4'], llk[subset(sum.patient.data, Status == 'P')$Patient,'4'])
```

#### Kmeans
```{r kmeansnorm, echo=F, message=F, warning=F, eval=F, fig.height=6, fig.width=6} 
for (pt in sum.patient.data$Patient) {
    print(pt)
    status = subset(sum.patient.data, Patient == pt)$Status
    x1 = patient.data[[pt]]$norm.seg.vals
    rownames(x1) = apply(patient.data[[pt]]$seg.vals[,c(1:4)], 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

    fit = kmeans(x1, 3)
    print(autoplot(fit, data=x1, label=T) + labs(title=paste(pt, '(', status, ') normalized seg', sep='')))
    pander(patient.data[[pt]]$variable.regions[,(1:4)], caption='Variable regions selected by cutoffs')
}
```

@@

## Using the segmentation values
sum.patient.data$years = with(sum.patient.data, paste(start.year, end.year, sep='-'))
sim = sum.patient.data[,c('Status','years')]
rownames(sim) = sum.patient.data[,'Patient']
for (pt in sum.patient.data$Patient) {
  patient.data[[pt]]$info = arrange(patient.data[[pt]]$info, Endoscopy.Year, desc(Pathology))

@@

  sim[pt, 'ns'] = net.similarity(ac)
  sim[pt, 'ss'] = sum.similarity(ac)
  sim[pt, 'clusters'] = length(clusters(ac))
  sim[pt, 'max.sd'] = max(sapply(x1, sd))

@@

@@

# fit = glm(Status~clusters, sim, family=binomial(link='logit'))
# pander(fit, justify='left')
# 
# sim = sim[, c('Status', 'clusters', 'ns','ss')]
# actualMatrix = matrix(0,1000,2,dimnames=list(c(), c('prec','recall')))
# for (i in 1:nrow(actualMatrix)) {
#   trainingRows = sample(1:nrow(sim), nrow(sim)-5)
#   train = sim[trainingRows,]
#   test = sim[-trainingRows,]
#   
#   fit = glm( (as.integer(Status)-1)~clusters, train, family=binomial(link='logit'))
#   pred = predict(fit, test, se.fit=F) # probabilities -- of which outcome though?
#   
#   actual = test$Status == 'P'
#   predicted = pred>0
#   
#   actualMatrix[i,'prec'] = getPrecision( actual, predicted )
#   actualMatrix[i, 'recall'] = getRecall( actual, predicted )
# }
# 
# ggplot(as.data.frame(actualMatrix), aes(x=recall, y=prec)) + geom_point() + 
#   geom_hline(yintercept = 0.5, col='red', alpha=0.5) +
#   labs(title='Precision/Recall for Progressor/NP prediction')

rc = cor.test(sim$max.sd, sim$clusters)
```

#### Cellularity?

We needed to check if cellularity was driving the difference in clusters. Overall that does not appear to be the case if we use the maximum standard deviation across all samples in a patient as a proxy: r=`r rc$estimate`, p-value=`r signif(rc$p.value, 2)`

@@

simX = simX[, c('Status', 'clusters', 'ns','ss','n.var')]
noHGDPR = as.data.frame(matrix(0,1000,2,dimnames=list(c(), c('prec','recall'))))
for (i in 1:nrow(actualMatrix)) {
  trainingRows = sample(1:nrow(simX), nrow(simX)-5)
  train = simX[trainingRows,]
  test = simX[-trainingRows,]

@@

#ggplot(as.data.frame(noHGDPR), aes(x=recall, y=prec)) + geom_point() + 
#  geom_hline(yintercept = 0.5, col='red', alpha=0.5) +
#  geom_vline(xintercept = 0.5, col='red', alpha=0.5) +
#  labs(title='Precision/Recall for Progressor/NP prediction NO HGD')

@@

ggplot(m, aes(x=factor(qv), y=value, col=variable)) + geom_point() +
  geom_errorbar(aes(ymin=value-sd, ymax=value+sd), width=.02) +
  geom_hline(yintercept=5, alpha=0.3) + 
  geom_hline(yintercept=15, alpha=0.3, col='red') + 
  scale_y_continuous(breaks=seq(0,max(m$value)+5, 10),labels=seq(0,max(m$value)+5, 10)  ) +
  labs(title='Clusters vs q parameters', x='q parameter', y='mean(clusters)')

pander(clus[,c('qv','NP','P')], caption="Mean of clusters by patient status per q parameter.")

ggplot(melt(pvals, measure.vars=c('p.clusters'), id.vars='qv'),aes(x=factor(qv),y=value, color=variable)) + 
  geom_point() + 
  labs(title='P vs NP p-values per q parameters', y='pvalue NP vs P', x='q param')
